Extending Graphical Representations for Compact Closed Categories with Applications to Symbolic Quantum Computation

### 2008-05-02 papers/reasoning-graphically/paper.tex
fixed a few typos.
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1474 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

A substantial strand of work in quantum computation is
the development of high-level models of quantum processes based on
compact closed  categories.  In this formalism, initiated in
\cite{AbrCoe:CatSemQuant:2004}, quantum processes---such as quantum
logic gates, or the measurement of a qubit---correspond to arrows
in a category, while the different quantum data types, usually just
arrays of qubits, are the objects.  The graphical presentation
described in the the preceding section thus provides a very expressive
notation for representing quantum processes.

@@

Coecke and Duncan \cite{Coecke:2008jo} propose a formal algebraic
system for quantum computation built from the following collection of
generators:  
\begin{description}
\item[Objects:] a single object, written $Q$.
\item[Arrows:] there are families of arrows $X$ and $Z$:
  \begin{align*}
  &\epsilon_Z : Q \to I, &\qquad&  \epsilon_X : Q \to I,\\
  &\delta_Z : Q \to Q \otimes Q, && \delta_X : Q \to Q \otimes Q, \\
  &\alpha_Z : Q \to Q &&  \alpha_X : Q \to Q   
  \end{align*}
  where $\alpha \in [0,2\pi)$, and in addition $H:Q\to Q$.
\end{description}
To each each arrow $f : A\to B$ we assign a formal adjoint $f^\dag : B
\to A$.  Each arrow is represented as a small graph; its adjoint is
the same graph written upside down.  We use colours to denote two families:
\begin{gather*}
  \epsilon_Z = \inlinegraphic{1.5em}{epsilon} \qquad
  \delta_Z = \inlinegraphic{1.5em}{delta} \qquad
  \epsilon_Z^\dag = \inlinegraphic{1.5em}{epsilondag} \qquad
  \delta_Z^\dag = \inlinegraphic{1.5em}{deltadag} \qquad
  \alpha_Z = \inlinegraphic{1.5em}{greenalpha} \qquad
\\  
  \epsilon_X = \inlinegraphic{1.5em}{redepsilon} \qquad
  \delta_X = \inlinegraphic{1.5em}{reddelta} \qquad
  \epsilon_X^\dag = \inlinegraphic{1.5em}{redepsilondag} \qquad
  \delta_X^\dag = \inlinegraphic{1.5em}{reddeltadag} \qquad
  \alpha_X = \inlinegraphic{1.5em}{redalpha} \qquad
\end{gather*}
The $H$ arrow is denoted by \inlinegraphic{1.5em}{H} and represents a
Hadamard gate. (The adjoints for $H$, $\alpha_X$ and
$\alpha_Z$ will be defined equationally.)  The free compact closed
category is then given by all graphs formed by composing and tensoring
these basic graphs.

Space does not permit a full exposition of this system, however we
can outline the basic intuitions behind these generators.  Each edge
in the picture represents a qubit, although several edges may
represent the same physical qubit at different times.  An edge may
even represent a ``virtual'' qubit which stands for a correlation
between different parts of the  system.  The maps
$\delta_Z$ and $\epsilon_Z$ represent quantum operations which
respectively copy and delete the eigenstates
of the  Pauli $Z$ operator.\footnote{Uniform copying operations are
  forbidden by the no-cloning theorem \cite{no-cloning}, but such
  operations are possible if we demand only the eigenstates of some
  self-adjoint operator to be copied. Other states will not not
  copied.  The same remarks hold true for erasing.}  
Notice that $\delta_Z$ has one edge
in its domain representing the qubit to copied, and two edges in its
codomain for the two copies it produces.  Similarly, $\epsilon_Z$ has
one qubit as input and no outputs.
The adjoints $\delta^\dag_Z$ and  $\epsilon^\dag_Z$ correspond to an
operation known as \emph{fusion}, and to the operation of preparing a
fresh qubit in a certain state. The $\alpha_Z$ corresponds to phase
shift of angle $\alpha$ in the $Z$ direction.  The family of maps
indexed by $X$ are defined in exactly the same way, but relative to
the Pauli $X$ rather than the $Z$.  All quantum operations may be
defined by combining these simple operations---which are essentially
classical---on two complementary observables.

\begin{remark}
  We emphasise that this is a mathematical notation is for
  representing quantum processes, unlike the so-called graph states
  used in measurement based quantum computation \cite{Briegel etc}
  which is based on certain quantum states.  Their graphical notation
  is dual to ours, with the vertices representing qubits and the edges
  representing entanglement between the individual quibits.  Within
  the formalism of this paper, a state is simply a process with no
  inputs; that  is, a concrete graph with empty domain.  An important
  application is to show that states or computations implmented
  differently are equivalent:  this includes graph states, and
  measurement based computations more generally. 
\end{remark}

@@

Consider the equations shown in Figure \ref{fig:graph-quant-eqns} which
involve only one colour:  these allow the remarkable \emph{spider theorem},
first noted in \cite{Coecke2006Quantum-Measure}, to be proved:

@@

Hence any connected subgraph involving nodes of only one colour,
wherever it occurs in a graph, may be collapsed to a single vertex
labelled by a single value $\alpha$, giving a ``spider''. Informally,
this can be depicted graphically as the equation in
Figure~\ref{fig:spider}-left. Conversely, a spider may be
arbitrarily divided into sub-spiders, provided the total in- and
out-degree is preserved, along with the sum of the $\alpha$s.
Further, one can derive, from the spider theorem, $n$-fold versions of
many of the other equations.

>>>>

A substantial strand of work in quantum computation has involved the
development of high-level models of quantum processes based on compact
closed categories.  In these formalisms, initiated in
\cite{AbrCoe:CatSemQuant:2004}, quantum processes---such as quantum
logic gates, or the measurement of a qubit---correspond to arrows in a
category, while the different quantum data types, usually just arrays
of qubits, are the objects.  The graphical representation described in
the preceding section thus provides a very expressive notation quantum
processes.

@@

Coecke and Duncan \cite{Coecke:2008jo} propose a formal algebraic
system for quantum computation built from the following collection of
generators:  
\begin{description}
\item[Objects:] a single object, written $Q$.
\item[Arrows:] there are families of arrows $X$ and $Z$:
  \begin{align*}
  &\epsilon_Z : Q \to I, &\qquad&  \epsilon_X : Q \to I,\\
  &\delta_Z : Q \to Q \otimes Q, && \delta_X : Q \to Q \otimes Q, \\
  &\alpha_Z : Q \to Q &&  \alpha_X : Q \to Q   
  \end{align*}
  where $\alpha \in [0,2\pi)$, and in addition $H:Q\to Q$.
\end{description}
To each each arrow $f : A\to B$ we assign a formal adjoint $f^\dag : B
\to A$.  Each arrow is represented as a small graph; its adjoint is
the same graph written upside down by reflection in the x-axis.  We
use colours to denote two families:
\begin{gather*}
  \epsilon_Z = \inlinegraphic{1.5em}{epsilon} \qquad
  \delta_Z = \inlinegraphic{1.5em}{delta} \qquad
  \epsilon_Z^\dag = \inlinegraphic{1.5em}{epsilondag} \qquad
  \delta_Z^\dag = \inlinegraphic{1.5em}{deltadag} \qquad
  \alpha_Z = \inlinegraphic{1.5em}{greenalpha} \qquad
\\  
  \epsilon_X = \inlinegraphic{1.5em}{redepsilon} \qquad
  \delta_X = \inlinegraphic{1.5em}{reddelta} \qquad
  \epsilon_X^\dag = \inlinegraphic{1.5em}{redepsilondag} \qquad
  \delta_X^\dag = \inlinegraphic{1.5em}{reddeltadag} \qquad
  \alpha_X = \inlinegraphic{1.5em}{redalpha} \qquad
\end{gather*}
The $H$ arrow is denoted by \inlinegraphic{1.5em}{H} and represents a
Hadamard gate. The adjoints for $H$, $\alpha_X$ and
$\alpha_Z$ will be defined equationally.  The free compact closed
category is then given by all graphs formed by composing and tensoring
these basic graphs.

In terms of quantum processes, each edge in a graph represents a
qubit, although several edges may represent the same physical qubit at
different times.  An edge may even represent a ``virtual'' qubit which
stands for a correlation between different parts of the system.  The
maps $\delta_Z$ and $\epsilon_Z$ represent quantum operations which
respectively copy and delete the eigenstates of the Pauli $Z$
operator.\footnote{Uniform copying operations are forbidden by the
  no-cloning theorem \cite{no-cloning}, but such operations are
  possible if we demand only the eigenstates of some self-adjoint
  operator to be copied.  Other states will not not copied.  The same
  remarks hold true for erasing.}  Notice that $\delta_Z$ has one edge
in its domain representing the qubit to be copied, and two edges in
its codomain for the two copies it produces.  Similarly, $\epsilon_Z$
has one qubit as input and no outputs.  The adjoints $\delta^\dag_Z$
and $\epsilon^\dag_Z$ correspond to an operation known as
\emph{fusion}, and to the operation of preparing a fresh qubit in a
certain state.  The $\alpha_Z$ corresponds to phase shift of angle
$\alpha$ in the $Z$ direction.  The family of maps indexed by $X$ are
defined in exactly the same way, but relative to the Pauli $X$
operator rather than the $Z$.  All quantum operations may be defined
by combining these simple operations---which are essentially
classical---on two complementary observables.

\begin{remark}
  We emphasise that this is a mathematical notation is for
  representing quantum processes, unlike the so-called graph states
  used in measurement based quantum computation \cite{Briegel etc}.
  Their graphical notation is dual to ours, with the vertices
  representing qubits and the edges representing entanglement between
  the individual quibits.  Within the formalism of this paper, a state
  is simply a process with no inputs; that is, a concrete graph with
  empty domain.  An important application is to show that states or
  computations implemented differently are equivalent: this includes
  graph states, and measurement based computations more generally.
\end{remark}

@@

Consider the equations from Figure \ref{fig:graph-quant-eqns} which
involve only one colour: these allow the remarkable \emph{spider
  theorem}, first noted in \cite{Coecke2006Quantum-Measure}, to be
proved:

@@

Hence any connected subgraph involving nodes of only one colour may be
collapsed to a single vertex, with a single value $\alpha$, giving a
``spider''. Informally, this can be depicted graphically as the
equation in Figure~\ref{fig:spider}-left. Conversely, a spider may be
arbitrarily divided into sub-spiders, provided the total in- and
out-degree is preserved, along with the sum of the $\alpha$s.
Furthermore, one can derive, from the spider theorem, $n$-fold versions of
many of the other equations.
